---
title: "Class9Candy_Project"
author: Gavin Ambrose PID A18548522
format: pdf
toc: true
---

## Importing candy data

First things first, let’s get the data from the FiveThirtyEight GitHub repo

```{r}
candy_file <- read.csv("https://raw.githubusercontent.com/fivethirtyeight/data/master/candy-power-ranking/candy-data.csv", row.names = 1)

head(candy_file)
```

## What is in the dataset

The dataset includes all sorts of information about different kinds of candy. For example, is a candy chocolaty? Does it have nougat? How does its cost compare to other candies? How many people prefer one candy over another?

> Q1. How many different candy types are in this dataset?

```{r}
str(candy_file)
```

There are 85 different candy types within this dataset


> Q2. How many fruity candy types are in the dataset?

```{r}
sum(candy_file$fruity)
```

There are 38 fruity candy types in the dataset

## What is your favorite candy?

One of the most interesting variables in the dataset is `winpercent`. For a given candy this value is the percentage of people who prefer this candy over another randomly chosen candy from the dataset. Higher values indicate a more popular candy.

> Q3. What is your favorite candy (other than Twix) in the dataset and what is it’s winpercent value?

```{r}
candy_file["Mike & Ike",]$winpercent
```

My favorite candy is Mike and Ike's and they have a winpercent value of 46.411

> Q4. What is the winpercent value for “Kit Kat”?

```{r}
candy_file["Kit Kat",]$winpercent
```

Kit Kat's have a winpercent of 76.768

> Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy_file["Tootsie Roll Snack Bars",]$winpercent
```

Tootsie Roll Snack Bars have a winpercent of 49.653

```{r}
library("skimr")
skim(candy_file)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent column is on a different scale than the majority of the data because it has a mean data of 50 whereas the others are between 0.08 and 0.4 means.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

I think that a one represents a belonging to that category wheras a 0 represents a lack of belonging to that category

## Exploratory analysis

A good place to start any exploratory analysis is with a histogram.

> Q8. Plot a histogram of winpercent values

```{r}
library(ggplot2)
ggplot(candy_file) + aes(winpercent) + geom_histogram() + stat_bin (bins = 10)
```


> Q9. Is the distribution of winpercent values symmetrical?

The histogram is sligthly skewed to the right. Making it assymmetric

> Q10. Is the center of the distribution above or below 50%?

The center of the distributuion is below 50%

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

```{r}
mean_chocolate <- mean(candy_file$winpercent[as.logical(candy_file$chocolate)])
mean_chocolate
mean_fruity <- mean(candy_file$winpercent[as.logical(candy_file$fruity)])
mean_fruity
```

Chocolate candy has a higher rating than fruity on average

> Q12. Is this difference statistically significant?

```{r}
t.test(candy_file$winpercent[as.logical(candy_file$chocolate)], candy_file$winpercent[as.logical(candy_file$fruity)])
```

We can reject the null hypothesis that there is a difference. The difference is significant because a p=2.871e-08 < 0.05

## Overall Candy Rankings

Let’s use the base R order() function together with `head()` to sort the whole dataset by winpercent. Or if you have been getting into the tidyverse and the dplyr package you can use the `arrange()` function together with `head()` to do the same thing and answer the following questions:

> Q13. What are the five least liked candy types in this set?

```{r}
library(dplyr)
candy_file |> arrange(winpercent) |> head(5)
```

The five least liked candies are Nik L Nip, Boston Baked Beans, Chiclets, Super Bubble, and Jawbusters.

> Q14. What are the top 5 all time favorite candy types out of this set?

```{r}
candy_file |> arrange(winpercent) |> tail(5)
```

The top 5 of all time are Snickers, Kit Kat, Reese's Minatures, and Reese's Peanut Butter cup.

> Q15. Make a first barplot of candy ranking based on winpercent values.

```{r}
ggplot(candy_file) + aes(winpercent, rownames(candy_file)) + geom_col()
```

> Q16. This is quite ugly, use the reorder() function to get the bars sorted by winpercent?

```{r}
ggplot(candy_file) + aes(winpercent, reorder(rownames(candy_file), winpercent)) + geom_col()
```

## Time to add some useful color

Let’s setup a color vector (that signifies candy type) that we can then use for some future plots. We start by making a vector of all black values (one for each candy). Then we overwrite chocolate (for chocolate candy), brown (for candy bars) and red (for fruity candy) values.

```{r}
my_cols=rep("black", nrow(candy_file))
my_cols[as.logical(candy_file$chocolate)] = "chocolate"
my_cols[as.logical(candy_file$bar)] = "brown"
my_cols[as.logical(candy_file$fruity)] = "pink"
```

```{r}
ggplot(candy_file) + aes(winpercent, reorder(rownames(candy_file), winpercent)) + geom_col(fill = my_cols)
```

> Q17. What is the worst ranked chocolate candy?

Sixlets

> Q18. What is the best ranked fruity candy?

Starbursts

## Taking a look at pricepercent

What about value for money? What is the best candy for the least money?

```{r}
library(ggrepel)

# How about a plot of win vs price
ggplot(candy_file) +
  aes(winpercent, pricepercent, label=rownames(candy_file)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

> Q19. Which candy type is the highest ranked in terms of `winpercent` for the least money - i.e. offers the most bang for your buck?

```{r}
ordleast <- order(candy_file$pricepercent, decreasing = TRUE)
head( candy_file[ordleast,c(11,12)], n=5 )
```


> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

```{r}
ordmost <- order(candy_file$pricepercent, decreasing = F)
head( candy_file[ordmost,c(11,12)], n=5 )
```

> Q21. Make a barplot again with geom_col() this time using pricepercent and then improve this step by step, first ordering the x-axis by value and finally making a so called “dot chat” or “lollipop” chart by swapping geom_col() for geom_point() + geom_segment().

```{r}
ggplot(candy_file) + aes(pricepercent, reorder(rownames(candy_file), pricepercent)) + geom_col()
```

## Make a lollipop chart of pricepercent

```{r}
ggplot(candy_file) +
  aes(pricepercent, reorder(rownames(candy_file), pricepercent)) +
  geom_segment(aes(yend = reorder(rownames(candy_file), pricepercent), 
                   xend = 0), col="gray40") +
    geom_point()
```

> One of the most interesting aspects of this chart is that a lot of the candies share the same ranking, so it looks like quite a few of them are the same price.

## Exploring the correlation structure

Now that we’ve explored the dataset a little, we’ll see how the variables interact with one another. We’ll use correlation and view the results with the corrplot package to plot a correlation matrix.

```{r}
library(corrplot)
corcandy <- cor(candy_file)
corrplot(corcandy)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

The two variables that have a strong negative correlation chocolate and fruity

> Q23. Similarly, what two variables are most positively correlated?

The chocolate variable and the winpercent variable have the strongest positive correlation.

## Principal Component Analysis

Let’s apply PCA using the prcomp() function to our candy dataset remembering to set the scale=TRUE argument.

```{r}
pcacandy <- prcomp(candy_file, scale = T)
summary(pcacandy)
```

Let's plot our main PCA scores (PCA1 and PCA2)

```{r}
plot(pcacandy$x[,1:2])
```

We can change the colors and characters

```{r}
plot(pcacandy$x[,1:2], col = my_cols, pch = 16)
```

LEt's try it with ggplot

```{r}
my_data <- cbind(candy_file, pcacandy$x[,1:3])
pcaggplot <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)
pcaggplot
```

Lets use greppel and label the plot

```{r}
pcaggplot + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

> Q24. Complete the code to generate the loadings plot above. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you? Where did you see this relationship highlighted previously?

```{r}
ggplot(pcacandy$rotation) +
  aes(x = PC1, y = reorder(rownames(pcacandy$rotation), PC1)) +
  geom_col()
```

The fruity, pluribis, and hard variables are picked up strongly by the PCA. These make sense because they are popular charicteristics of candy. This was highlighted in the previous question that also use bar plots.

## Summary

> Q25. Based on your exploratory analysis, correlation findings, and PCA results, what combination of characteristics appears to make a “winning” candy? How do these different analyses (visualization, correlation, PCA) support or complement each other in reaching this conclusion?

Based on visual data, chocolate and bar candies seems to preform the best, and fruity candies the worst. Positive correlation is associated with chocolate and bar candies, while a negative correlation is associated with fruity candies. PC1 has a shows chocolate and bar variables along one of the axis and fruity and plurabis along a different axis. These all propose that the "winning" candy is likely a chocolate candy in a bar form.











