---
title: "Class 5: Data Vis with ggplot"
author: "Gavin Ambrose (PID: A18548522)"
format: pdf
toc: true
---

## Background

There are lots of ways to make figures in R. These include so-called "base R" graphics (e.g. `plot()`) and tons of add-on packages like **ggplot2**.

For example, here we make the same plot with both:

```{r}
head(cars)

plot(cars)
```

First, I need to install the package with the command `install.packages()`.

>**NB** We never run an installl cmd in a quartro code chunk or we will end up re-installing packages many, many times - which is not what we want!

Every time we want to use one of these "add-ons" packages, we need to load it up with the `library()` function.

```{r}
library(ggplot2)

ggplot(cars)
```

Every ggplot needs at least 3 things:

- The **data**, the stuff you want plotted
- The **aes**thitics, , how the data map to the plot
- The **geom**etry, the type of plot

Add a line to better show relationship between speed and dist wtih `geom_smooth()` and setting `method = "lm"`

```{r}
P <- ggplot(cars) +
  aes(x = speed, y = dist) +
  geom_point() +
  geom_smooth(method = "lm", se= FALSE) +
  labs(title = "Stopping Distance of Old Cars", 
       subtitle = "Data from the `cars` object",
       x = "Speed (mph)",
       y = "Distance (ft)")

```

Titles can be added using the `labs()` function


By saving it as a variable, it doesn't print the chart, let's render it out

```{r}
P
```

```{r}
P + theme_minimal()
```

By using `theme_minimal`, it removes the the gray background from the ggplot

The `plot()` cmd is *less code*, but gives *less formatting options*. The `ggplot()` cmd is *more code*, but has *more formatting options* available

## Gene expression plot

We can read the input data from the class website!

```{r}
url <- "https://bioboot.github.io/bimm143_S20/class-material/up_down_expression.txt"
genes <- read.delim(url)
head(genes)
```

A first version plot

```{r}
ggplot(genes,) +
  aes(Condition1, Condition2) +
  geom_point()
```


```{r}
table(genes$State)
```

Version 2, let's color by `State` so we can see the "up" and "down" significant genes compared to all the "unchanged" genes

```{r}
ggplot(genes,) +
  aes(Condition1, Condition2, col = State) +
  geom_point()
```

Version 3, let's modify the default colors to something we like.

```{r}
ggplot(genes,) +
  aes(Condition1, Condition2, col = State) +
  geom_point() +
  scale_color_manual(values = c("darkblue", "gray50", "darkred")) + labs(x = "Control (No Drug)",y = "No Drug", title = "Gene Expression Changes upon GLP-1 Drug") +
  theme_minimal()
```

## Going Further

Let's have a look at the famous **gapminder** dataset

```{r}
url <- "https://raw.githubusercontent.com/jennybc/gapminder/master/inst/extdata/gapminder.tsv"

gapminder <- read.delim(url)

head(gapminder, 3)
```

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col = continent) +
geom_point(alpha = 0.5)
```

Let's "facet" (i.e. make a separate plot) by continent rather than the big hot mess above.

```{r}
ggplot(gapminder) +
  aes(gdpPercap, lifeExp, col = continent) +
geom_point(alpha = 0.5) + 
  facet_wrap(~continent) + 
  labs(x = "GDP per Capita", y = "Life Expectancy", title = "Life Expectancy based on GDP per Capita", subtitle = "Graph Seperated by Continent")
```

## Custom Plot

How big big is this gapmider data set
Use `nrow` to specify the number of rows in a data set and `ncol` to specify the number of columns in a dataset

```{r}
nrow(gapminder)
ncol(gapminder)
```

I want to "filter" down to a subset of this data. I will use the **dplyr** package to help me.

First I need to install it and then load it up... `install.packages("dplyr")` and then `library(dplyr)`

```{r}
library(dplyr)
gap2007 <- filter(gapminder, year ==2007)
head(gap2007)
```
What is the life expectancy of United States and Ireland in 2007?

```{r}
filter(gap2007, country == "Ireland")
```
 or 

```{r}
filter(gapminder, year == 2007, country == "Ireland")
```

What is the life expectancy in the United states in 2007?

```{r}
filter(gapminder, year == 2007, country == "United States")
```
> Q. Make a plot comparing 1977 and 2007 for all countries

```{r}
filter_data <- filter(gapminder, year %in% c(1977, 2007))
ggplot(filter_data) + 
  aes(gdpPercap, lifeExp, col = continent) + 
  geom_point(alpha = 0.6) + 
  facet_wrap(~year) +
  labs(x = "GDP per Capita", y = "Life Expectancy", title = "Life Expectancy based on GDP per Capita", subtitle = "Graphs seperated by Year")
```




