---
title: "Class 18: Investigating Pertussis"
author: "Gavin Ambrose PID: A18548522"
format: pdf
toc: true
---

## Background

Pertussis (more commonly known as whooping cough) is a highly contagious respiratory disease caused by the bacterium Bordetella pertussis.

People of all ages can be infected leading to violent coughing fits followed by a characteristic high-pitched “whoop” like intake of breath. Children have the highest risk for severe complications and death.

## 1. Investigating pertussis cases by year

Let's read the information from the following website:
<https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html>

> Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

```{r, message=FALSE}
library(ggplot2)
library(datapasta)
```

Here is the data

```{r, echo=FALSE}
cdc <- data.frame(
  year = c(
    1922L,
    1923L,
    1924L,
    1925L,
    1926L,
    1927L,
    1928L,
    1929L,
    1930L,
    1931L,
    1932L,
    1933L,
    1934L,
    1935L,
    1936L,
    1937L,
    1938L,
    1939L,
    1940L,
    1941L,
    1942L,
    1943L,
    1944L,
    1945L,
    1946L,
    1947L,
    1948L,
    1949L,
    1950L,
    1951L,
    1952L,
    1953L,
    1954L,
    1955L,
    1956L,
    1957L,
    1958L,
    1959L,
    1960L,
    1961L,
    1962L,
    1963L,
    1964L,
    1965L,
    1966L,
    1967L,
    1968L,
    1969L,
    1970L,
    1971L,
    1972L,
    1973L,
    1974L,
    1975L,
    1976L,
    1977L,
    1978L,
    1979L,
    1980L,
    1981L,
    1982L,
    1983L,
    1984L,
    1985L,
    1986L,
    1987L,
    1988L,
    1989L,
    1990L,
    1991L,
    1992L,
    1993L,
    1994L,
    1995L,
    1996L,
    1997L,
    1998L,
    1999L,
    2000L,
    2001L,
    2002L,
    2003L,
    2004L,
    2005L,
    2006L,
    2007L,
    2008L,
    2009L,
    2010L,
    2011L,
    2012L,
    2013L,
    2014L,
    2015L,
    2016L,
    2017L,
    2018L,
    2019L,
    2020L,
    2021L,
    2022L,
    2023L,
    2024L,
    2025L
  ),
  cases = c(
    107473,
    164191,
    165418,
    152003,
    202210,
    181411,
    161799,
    197371,
    166914,
    172559,
    215343,
    179135,
    265269,
    180518,
    147237,
    214652,
    227319,
    103188,
    183866,
    222202,
    191383,
    191890,
    109873,
    133792,
    109860,
    156517,
    74715,
    69479,
    120718,
    68687,
    45030,
    37129,
    60886,
    62786,
    31732,
    28295,
    32148,
    40005,
    14809,
    11468,
    17749,
    17135,
    13005,
    6799,
    7717,
    9718,
    4810,
    3285,
    4249,
    3036,
    3287,
    1759,
    2402,
    1738,
    1010,
    2177,
    2063,
    1623,
    1730,
    1248,
    1895,
    2463,
    2276,
    3589,
    4195,
    2823,
    3450,
    4157,
    4570,
    2719,
    4083,
    6586,
    4617,
    5137,
    7796,
    6564,
    7405,
    7298,
    7867,
    7580,
    9771,
    11647,
    25827,
    25616,
    15632,
    10454,
    13278,
    16858,
    27550,
    18719,
    48277,
    28639,
    32971,
    20762,
    17972,
    18975,
    15609,
    18617,
    6124,
    2116,
    3044,
    7063,
    22538,
    21996
  )
)
```

Let's creat the ggplot.

```{r}
ggplot(cdc) + 
  aes(year, cases) + 
  geom_point() + 
  geom_line() +
  labs(x = "Years", y ="Number of Cases", title = "Number of Cases Over Time" ) +
  theme_minimal()
```

> Q. Add the annotation lines for major milestones of wP vaccination roll-out (1946) and the switch to the ap vaccine (1996)

## 2. A tale of two vaccines (wP & aP)

```{r}
ggplot(cdc) + 
  aes(year, cases) + 
  geom_point() + 
  geom_line() +
  labs(x = "Years", y ="Number of Cases", title = "Number of Cases Over Time" ) +
  theme_minimal() +
geom_vline(xintercept = 1946, col = "blue", lty = 2) +
  geom_vline(xintercept = 1996, col = "red", lty = 2) +
  geom_vline(xintercept = 2020, col = "gray", lty = 2)
```
> Q2. Using the ggplot geom_vline() function add lines to your previous plot for the 1946 introduction of the wP vaccine and the 1996 switch to aP vaccine (see example in the hint below). What do you notice?

The number of cases decreases initially by oscillating downwards before leveling out

> Q3. Describe what happened after the introduction of the aP vaccine? Do you have a possible explanation for the observed trend?

There is a general increase in the number of cases. This could be due to more sensitive testing that has been developed, the increased resistance of the bacteria, or a reduced effectiveness of the drug (concentration)

## 3. Exploring CMI-PB data

Why is this vaccine-preventable disease on the upswing? To answer this question we need to investigate the mechanisms underlying waning protection against pertussis. This requires evaluation of pertussis-specific immune responses over time in wP and aP vaccinated individuals.

```{r, message=FALSE}
library(jsonlite)
```

Let’s now read the main subject database table from the CMI-PB API. You can find out more about the content and format of this and other tables here: <https://www.cmi-pb.org/blog/understand-data/>

To read these types of files into R we will use the `read_json()` function from the jsonlite package

```{r}
subject <- read_json("https://www.cmi-pb.org/api/subject",
                     simplifyVector = T) 
head(subject, 3)
```

> Q4. How many aP and wP infancy vaccinated subjects are in the dataset?

```{r}
table(subject$infancy_vac)
```

There are 58 infants vaccinated with the wP vaccine and 60 vaccinated with the aP vaccine.

> Q5. How many Male and Female subjects/patients are in the dataset?

```{r}
table(subject$biological_sex)
```

There are 112 female patients in the dataset and 60 males within the dataset

> Q6. What is the breakdown of `race` and `biological sex` (e.g. number of Asian females, White males etc…)?

```{r}
table(subject$race, subject$biological_sex)
```

There is a large poportion of the sample that contains Asian and White population with little sampling from American Indian/Alaska Native, Black or African American, Native Hawaiian or Other Pacific Islander, with moderate reporting form More than one race or Unknown or Not reported

> Q. Is this representative?

No, this data pulls unequally from the different races and genders.

### Working with dates

> Q7. Using this approach determine (i) the average age of wP individuals, (ii) the average age of aP individuals; and (iii) are they significantly different?

```{r}

```

### Joining multiple tables

Let's read some more tables

```{r}
specimen <- read_json("http://cmi-pb.org/api/v5_1/specimen", 
                      simplifyVector = T )
ab_titer <- read_json("http://cmi-pb.org/api/v5_1/plasma_ab_titer",
                      simplifyVector = T)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```

To know whether a given `specimen_id` comes from an aP or wP individual we need to link (a.k.a. “join” or "merge") our `specimen` and `subject` data frames. Only then will we know that a given AB measurement (from the `ab-tier` table) was collected on a certain data from a certain date (from the `specimen` table) from the certain wP or aP subject (from the `subject` table).

We can use **dplyr** and the `*join()` family of functions to do this.

> Q9. Complete the code to join specimen and subject tables to make a new merged data frame containing all specimen records along with their associated subject details:

```{r,message=FALSE}
library(dplyr)
```

```{r}
meta <- inner_join(subject, specimen)
head(meta)
```

> Q10. Now using the same procedure join meta with titer data so we can further analyze this data in terms of time of visit aP/wP, male/female etc.

```{r}
abdata <- inner_join(ab_titer, meta)
head(abdata)
```

> Q11. How many specimens (i.e. entries in abdata) do we have for each isotype?

```{r}
table(abdata$isotype)
```

There are several thousand entries for each isotypes available. There are 6698 for IgE, 7265 fir IgG, 11993 for IgG1, 12000 for IgG2, 12000 for IgG3, and 12000 for IgG4.

> Q. How many different "antigen" values are measured

```{r}
table(abdata$antigen)
```

> Q12. What are the different $dataset values in abdata and what do you notice about the number of rows for the most “recent” dataset?

```{r}
table(abdata$dataset)
```

There are quite a few different types of antigens and each contain varying amounts of values associated with them

## Examine IgG Ab titer levels

Let's focus on IgG isotype 

```{r}
igg <- abdata |> 
  filter(isotype == "IgG")
head(igg)
```

> Q13. Complete the following code to make a summary boxplot of Ab titer levels (MFI) for all antigens: 

Make a plot of `MFI_normalized` vs `antigen`

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen) +
  geom_boxplot() +
  theme_bw()
```

> Q14. What antigens show differences in the level of IgG antibody titers recognizing them over time? Why these and not others?

The antigens "PT", "FIM2/3", and "FHA" apppear to have the widest range of values

> Q. Is there a difference for these responses between aP and wP individuals?

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  theme_bw()
```

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~infancy_vac) +
  theme_bw()
```

> Q. Is there a difference with time (i.e. before booster shot vs after booster shot)

```{r}
ggplot(igg) +
  aes(MFI_normalised, antigen, col=infancy_vac) +
  geom_boxplot() +
  facet_wrap(~visit) +
  theme_bw()
```

Let's do one last plot:

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset") # Filtering per year

abdata.21 %>% 
  filter(isotype == "IgG",  antigen == "PT") %>% #Looking at specific isotype
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) + #draw lines for each indicudual patient
    geom_point() +
    geom_line() +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)") +
  theme_bw()
```

> Q. Can you add a `geom_smooth()` to show the difference between aP and wP

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset") # Filtering per year

abdata.21 %>% 
  filter(isotype == "IgG",
         antigen == "PT") %>% #Looking at specific isotype
  ggplot() +
    aes(x=planned_day_relative_to_boost,
        y=MFI_normalised,
        col=infancy_vac,
        group=subject_id) + #draw lines for each indicudual patient
    geom_point(alpha = 0.3) +
    geom_line(alpha = 0.3) +
    geom_vline(xintercept=0, linetype="dashed") +
    geom_vline(xintercept=14, linetype="dashed") +
  labs(title="2021 dataset IgG PT",
       subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)") +
  theme_bw() +
  geom_smooth(
  aes(group = infancy_vac),
  se = FALSE,
  span = 0.3
)
```













