---
title: "Class 17: Analyzing sequencing data"
author: "Gavin Ambrose PID: A18548522"
format: pdf
toc: true
toc-depth: 4
---

## Importing Data from Shell computer

With each sample having its own directory containing the Kallisto output, we can import the transcript count estimates into R using:

```{r,message=FALSE}
library(tximport)
library(rhdf5)
```

```{r}
folders <- dir(pattern="SRR21568*")
samples <- sub("_quant", "", folders)
files <- file.path( folders, "abundance.h5" )
names(files) <- samples

txi.kallisto <- tximport(files, type = "kallisto", txOut = TRUE)
```

```{r}
head(txi.kallisto$counts)
```

```{r}
colSums(txi.kallisto$counts)
```

```{r}
sum(rowSums(txi.kallisto$counts)>0)
```

### Filtering Data

Before subsequent analysis, we might want to filter out those annotated transcripts with no reads:

```{r}
to.keep <- rowSums(txi.kallisto$counts) > 0
kset.nonzero <- txi.kallisto$counts[to.keep,]
```

```{r}
keep2 <- apply(kset.nonzero,1,sd)>0
x <- kset.nonzero[keep2,]
```

## PCA Ananlysis

We can now apply any exploratory analysis technique to this counts matrix. As an example, we will perform a PCA of the transcriptomic profiles of these samples.

```{r}
pca <- prcomp(t(x), scale=TRUE)
summary(pca)
```

### PCA Visualization

Now we can use the first two principal components as a co-ordinate system for visualizing the summarized transcriptomic profiles of each sample:

```{r}
plot(pca$x[,1], pca$x[,2],
     col=c("blue","blue","red","red"),
     xlab="PC1", ylab="PC2", pch=16)
```

### Making a ggplot version

```{r}
library(ggplot2)
```

Let's create a data frame for the data

```{r}
library(ggrepel)

# Make metadata object for the samples
colData <- data.frame(condition = factor(rep(c("control", "treatment"), each = 2)))
rownames(colData) <- colnames(txi.kallisto$counts)

# Make the data.frame for ggplot 
y <- as.data.frame(pca$x)
y$Condition <- as.factor(colData$condition)

```

```{r}
ggplot(y) +  
  aes(PC1,PC2,col = Condition) + 
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_minimal()
```

#### PC1 vs PC 3

```{r}
ggplot(y) +  
  aes(PC1,PC3,col = Condition) + 
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_minimal()
```

#### PC2 vs PC3

```{r}
ggplot(y) +  
  aes(PC2,PC3,col = Condition) + 
  geom_point() +
  geom_text_repel(label=rownames(y)) +
  theme_minimal()
```



